---
title: "2020 Ab titer data: Median (at Day 0) based normalisation"
output:
  html_document:
    df_print: paged
---

```{r echo=FALSE}
packages <- c("tidyverse", "hrbrthemes", "jsonlite")

for (n in 1:length(packages)) {
    suppressMessages(library(packages[n], character.only = TRUE))
}

```

### Reading dataset from experimental datafile
```{r}
#setwd("/home/pramod/Documents/cmipb_project/ab_plate_normalisation")
setwd("/home/pramod/Documents/GitHub/gitlab/2021-ab-titer-data-normalisation/private/data/")
titers_2020 <- read_tsv("2020_abtiter_long.csv", col_names = FALSE, show_col_types = FALSE) 
colnames(titers_2020) <- c("specimen_id",  "isotype",  "is_antigen_specific",  "antigen",  "ab_titer", "unit",  "lower_limit_of_detection")

titers_2020 <- titers_2020 %>%
  mutate(
    antigen = replace(antigen, antigen =='1% PFA PT', 'PT'),
    isotype_antigen = paste0(isotype,"_",antigen, "_", unit),
    ab_titer_original = ab_titer, ## Kepping ab)titer_original for final dataframe
    subject_id = substr(specimen_id, 1,4),
    visit = substr(specimen_id, 9,9),
    planned_day_relative_to_boost= case_when(
                visit == 1 ~ 0,
                visit == 2 ~ 1,
                visit == 3 ~ 3,
                visit == 4 ~ 7,
                visit == 5 ~ 14,
                visit == 6 ~ 30,
                visit == 7 ~ 90,
           )
  ) 
```

### Count NA and below LOD samples per Isotype and Antigen pair 
### Remove features with more than 80% missing values
```{r}
count_samples <- titers_2020 %>%
  group_by(isotype_antigen) %>%
  summarise(count_samples = n())

count_0 <- titers_2020 %>%
  group_by(isotype_antigen) %>%
  filter(is.na(ab_titer) == TRUE | ab_titer < lower_limit_of_detection) %>%
  summarise(count_0 = n(), percentage_0 = (count_0/394)* 100) %>%
  arrange(desc(percentage_0))
  
count_0

remove_features <- count_0[count_0$percentage_0 > 90,]$isotype_antigen

#titers_2020_feature_removed <- titers_2020[!titers_2020$isotype_antigen %in% remove_features, ]

for(select_ia in remove_features)
{
plot1 <- titers_2020 %>%
  mutate(subject_id = as.factor(subject_id)) %>%
  filter(isotype_antigen == select_ia, planned_day_relative_to_boost < 50) %>%
  arrange(desc(ab_titer)) %>%
  ggplot(aes(y=ab_titer)) +  stat_ecdf(geom = "point", pad = FALSE) +
  labs(y = "Ab titer", x = "Percentile")+ ggtitle(paste0("2020 Antibody titers (" , select_ia, ")")) + theme_bw() +
  scale_y_continuous(trans = 'log2', labels=scaleFUN)

plot(plot1)
}
```
### Replace not detected values with lower_limit_of_detection
```{r}
titers_2020_feature_removed = titers_2020
titers_2020_feature_removed_lod <- titers_2020_feature_removed %>%
  mutate(
    lower_limit_of_detection = if_else(isotype_antigen == "IgE_Total_UG/ML", 2.09613325980121, lower_limit_of_detection), ## Correct LOD for IgE_Total_UG/ML
    ab_titer = if_else(ab_titer < lower_limit_of_detection, lower_limit_of_detection, ab_titer)
  ) 

## QC for LOD outliers
## No Outliers detected
lod_outlier <- titers_2020_feature_removed_lod %>%
  group_by(isotype_antigen) %>%
  #slice_min(MFI_normalized, n=2) %>%
  #arrange(desc(MFI_normalized)) %>%
  summarise(min = min(ab_titer), lod = lower_limit_of_detection, fc_min_max = lower_limit_of_detection/min(ab_titer)) %>%
  filter(fc_min_max > 3)
```


### Identifying outlier ab titers.
```{r}
## If second_max_value/max_value > 3 then set max_value = LOD
top_outlier <- titers_2020_feature_removed_lod %>%
  group_by(isotype_antigen) %>%
  slice_max(ab_titer, n=2) %>%
  arrange(desc(ab_titer)) %>%
  summarise(fc_min_max = max(ab_titer)/min(ab_titer)) %>%
  filter(fc_min_max > 3)
  
## If first_value/LOD > 3 then set LOD as first_value
bottom_outlier <- titers_2020_feature_removed_lod %>%
  group_by(isotype_antigen) %>%
  slice_min(ab_titer, n=2) %>%
  arrange(desc(ab_titer)) %>%
  summarise(fc_min_max = max(ab_titer)/min(ab_titer)) %>%
  filter(fc_min_max > 2)

## Ploting cummulative distribution to identify outliers.
#Our transformation function
scaleFUN <- function(x) sprintf("%.3f", x)

#for(select_ia in c('IgE_Total'))
for(select_ia in unique(titers_2020_feature_removed_lod$isotype_antigen))
{
plot1 <- titers_2020_feature_removed_lod %>%
  mutate(subject_id = as.factor(subject_id)) %>%
  filter(isotype_antigen == select_ia, planned_day_relative_to_boost < 50) %>%
  arrange(desc(ab_titer)) %>%
  ggplot(aes(y=ab_titer)) +  stat_ecdf(geom = "point", pad = FALSE) +
  labs(y = "Ab titer", x = "Percentile")+ ggtitle(paste0("2020 Antibody titers (" , select_ia, ")")) + theme_bw() +
  scale_y_continuous(trans = 'log2', labels=scaleFUN)

plot(plot1)
}
```

### LOD Corrections
#### 1. Top outliers: IgG_PT
#### 2. Bottom outliers: IgG_FHA, IgG_PRN
```{r}
titers_2020_feature_removed_lod_corrected <- titers_2020_feature_removed_lod %>%
  mutate(
    lower_limit_of_detection = if_else(isotype_antigen == "IgG_FHA_IU/ML", 4.67953450834645, lower_limit_of_detection), ## Correct LOD for IgG_FHA
    lower_limit_of_detection = if_else(isotype_antigen == "IgG_PRN_IU/ML", 6.20594906363301, lower_limit_of_detection), ## Correct LOD for IgG_PRN
    ab_titer = if_else(isotype_antigen == "IgG_PT_IU/ML" & ab_titer > 15000 , lower_limit_of_detection, ab_titer), ## Correct top outlier for IgG_PT 
  ) %>%
  mutate(ab_titer = if_else(ab_titer < lower_limit_of_detection | is.na(ab_titer) == TRUE, lower_limit_of_detection, ab_titer)) ## Setting ab titers to LOD for new LOD definitions

## Validation plots
for(select_ia in c("IgG_FHA_IU/ML", "IgG_PRN_IU/ML", "IgG_PT_IU/ML"))
{
plot1 <- titers_2020_feature_removed_lod_corrected %>%
  mutate(subject_id = as.factor(subject_id)) %>%
  filter(isotype_antigen == select_ia, planned_day_relative_to_boost < 50) %>%
  arrange(desc(ab_titer)) %>%
  ggplot(aes(y=ab_titer)) +  stat_ecdf(geom = "point", pad = FALSE) +
  labs(y = "Ab titer", x = "Percentile")+ ggtitle(paste0("2020 Antibody titers (" , select_ia, ")")) + theme_bw() +
  scale_y_continuous(trans = 'log2', labels=scaleFUN)

plot(plot1)
}

```
### Performing median based normalisation

```{r}
## Setting MFI zero values to lower limit of detection
titers2020_calculate_lod <- titers_2020_feature_removed_lod_corrected %>%
  #filter(isotype_antigen == 'IgE_Total') %>%
  group_by(isotype_antigen) %>%
  mutate(
         lod_new = min(ab_titer[ab_titer > 0], na.rm = TRUE),
         MFI_na_removed = if_else(ab_titer == 0 | is.na(ab_titer) == TRUE, lod_new, ab_titer)
         ) %>%
  ungroup() 

## Calculate Overall median using MFI at day post boost 0
df_d0_median_2020  <- titers2020_calculate_lod %>%
  filter(planned_day_relative_to_boost == 0) %>%
  group_by(isotype_antigen) %>%
  summarise(
    MFI_median_d0 = median(MFI_na_removed),
  ) %>%
  ungroup() 

titers_2020_new_raw   <-  left_join(titers2020_calculate_lod , df_d0_median_2020 , by = "isotype_antigen") %>%
  mutate(
    MFI_normalized = MFI_na_removed / MFI_median_d0,
  )
```

### Plot longitudinal data after median based normalisation
```{r}

#for(select_ia in c('IgE_Total_IU/ML'))
for(select_ia in unique(titers_2020_new_raw$isotype_antigen))
{
plot1 <- titers_2020_new_raw %>%
  mutate(subject_id = as.character(subject_id)) %>%
  filter(isotype_antigen == select_ia, planned_day_relative_to_boost < 50) %>%
    ggplot(aes(x=planned_day_relative_to_boost, y=MFI_normalized, )) +
      geom_line(aes(group=subject_id),linetype = "dotted") +
      geom_point() + 
      labs(x = "Day post boost", y = "MFI Normalised") + 
      geom_smooth(size = 1) +
      theme_bw() +
      theme(strip.background = element_blank(), strip.placement = "outside") +
      ggtitle(paste0("2020 Antibody titers: ", select_ia)) +
      scale_y_continuous(trans = 'log2', labels=scaleFUN)

plot(plot1)
}
```
### Output/final dataframe for database porpose
```{r}
titers_2020_new <- titers_2020_new_raw %>%
  mutate(ab_titer_normalised = MFI_normalized,
         ab_titer = ab_titer_original) %>%
  select(specimen_id, isotype, antigen, unit, is_antigen_specific, ab_titer, ab_titer_normalised, lower_limit_of_detection) 

##write.table(titers_2020_new,"2020_abtiters_normalised_04052021.csv", row.names = FALSE, sep="\t")
```

